---
title: "DeepPAM Basics"
author: "Lukas"
date: "`r Sys.time()`"
output:
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    self_contained: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  message=FALSE, error=TRUE, warning=FALSE
)

# Modeling
library(keras)
library(mgcv)

# Convenience
library(dplyr)
library(broom)
library(ggplot2)

# Sacrificing a goat to the python env
reticulate::use_condaenv("r-reticulate", required = TRUE)
reticulate:::ensure_python_initialized()
tensorflow::tf_config()
```

# Linear Models

```{r lm-setup}
set.seed(17)
n <- 700
x1 <- runif(n, 0, 10)
x2 <- runif(n, -2, 2)
x3 <- sample(c(0, 1), n, replace = TRUE)

y <- 1 + 2 * x1 + 3 * x2 + 4 * x3 + rnorm(n, 0, 1)
df <- data.frame(y, x1, x2, x3)
```

## `lm()`

```{r lm-lm-fit}
lm_lm <- lm(y ~ x1 + x2 + x3, data = df)

tidy(lm_lm)
glance(lm_lm)
```

## `keras`

- Single layer NN
- Linear regression layer with bias

```{r lm-nn-fit, cache=TRUE}
lm_nn <- keras_model_sequential(name = "LM") %>%
  layer_dense(units = 1, activation = "linear", use_bias = TRUE, name = "Regression")

lm_nn %>%
  compile(
    loss = "mse",
    optimizer = optimizer_rmsprop(),
    metrics = NULL
  )

lm_nn %>%
  fit(
    as.matrix(df[, 2:4]), df$y,
    epochs = 50,
    batch_size = 8,
    validation_split = 0.2,
    verbose = TRUE
  )

lm_nn

get_weights(lm_nn)
```

Close enough?

# Generalised Additive Models

```{r gam-setup}
set.seed(17)
n <- 700
x1 <- runif(n, 0, 10)
x2 <- runif(n, -2, 2)
x3 <- sample(c(0, 1), n, replace = TRUE)

predictor <- -3 + 0.05 * x1^2 + x2 + x3
prob <- exp(predictor) / (1 + exp(predictor))
y <- rbinom(n, 1, prob)
df <- data.frame(y, x1, x2, x3)
```

## `mgcv()`

> Use the mgcv package to estimate a GAM with linear effects of x2 and x3 and a smooth x1 effect.

```{r gam-mgcv-fit}
gam_mgcv <- gam(y ~ s(x1, fx = FALSE) + x2 + x3, family = binomial(link = "logit"), data = df)

gam_mgcv$coefficients
# smoothing param for later
gam_mgcv$sp
# S matrix
gam_mgcv$smooth[[1]]$S[[1]]
```

Takeaway: I need to do some GAM-reading.

## `keras`

> Neglect the penalisation of the smooth eﬀect and construct an equivalent neural network with keras and tensorflow. Fit it.

- Two parallel layers for linear / smooth effects? 🤔
- [Relevant Keras docs maybe](https://keras.rstudio.com/articles/functional_api.html)

```{r gam-keras-fit}
# Define inputs, two linear + 1 smooth effect
linear_input <- layer_input(shape = c(2), name = "linear_in")
smooth_input <- layer_input(shape = c(9), name = "smooth_in")

# Define intermediate outputs
linear_out <- linear_input %>%
  layer_dense(units = 1, activation = "linear", name = "linear_effect")

# Until I figure out what to use for a smooth effect here
# mgcv shows 9 components of the smooth effects, but just 1 neuron though
smooth_out <- smooth_input %>%
  layer_dense(units = 1, activation = "linear", name = "smooth_effect", use_bias = FALSE)

# layer_concatenate sums up intermediate outputs
# sigmoid activation for binary outcome
combined_out <- layer_add(c(linear_out, smooth_out)) %>%
  layer_activation(activation = "sigmoid", name = "output_prob")

# Combine the above to a keras model
gam_nn <- keras_model(
  inputs = c(linear_input, smooth_input),
  outputs = c(combined_out)
)

gam_nn %>%
  compile(
    loss = "binary_crossentropy", # That's Bernoulli logLik right?
    optimizer = optimizer_rmsprop(),
    metrics = NULL
  )

gam_nn

gam_nn %>%
  fit(
    # First linear input (x2, x3), then smooth input (x1)
    x = list(as.matrix(df[, 3:4]), model.matrix(gam_mgcv)[, 4:12]),
    y = df$y,
    epochs = 200,
    batch_size = 32,
    validation_split = 0.2,
    verbose = TRUE
  )

get_weights(gam_nn)
```


## Plot comparion

> Use the plot function to plot the mgcv model’s smooth term.

```{r gam-mgcv-plot}
plot(gam_mgcv, select = 1, resid = TRUE, pch = 1, cex = 1, shade = TRUE)
```

> Create an equivalent plot for the neural network’s smooth term. You will need to predict over a grid of x1. Bring both plots together in a single plot.

```{r gam-keras-plot}
test_df <- tibble::tibble(
  y = rep(0, 100),
  x1 = seq(-10, 10, length.out = 100),
  x2 = rep(0, 100),
  x3 = rep(0, 100)
)

lpmat <- predict(gam_mgcv, newdata = test_df, type = "lpmatrix")
gam_nn_preds <- predict(gam_nn, x = list(as.matrix(lpmat[, 2:3]), lpmat[, 4:12]))

test_df %>%
  mutate(yhat = as.numeric(gam_nn_preds)) %>%
  ggplot(aes(x = x1, y = yhat)) + 
  geom_point() +
  geom_path() +
  theme_minimal()
```


## `keras` w/ Penalisation

> Read [here](https://m-clark.github.io/generalized-additive-models/technical.html) about the penalisation of smooth terms in mgcv. Search through the mgcv model for all components of the penalisation.

> Apply the proper penalisation to the equivalent neural network and refitt the network. Compare the resulting smooth terms with meaningful plots (see 3)). Use the extracted components from 4) You will find the code below helpful.

```r
kernel_regularizer = function(x) {
  k_mean(k_batch_dot(x, k_dot(
  tf$constant(???, dtype = "float64"), x), axes = 2))
}
```

From Wood 2017:

$$||\boldsymbol{y} - \boldsymbol{X}\boldsymbol{\beta}||^2 + \underbrace{\lambda \boldsymbol{\beta}^\top \boldsymbol{S} \boldsymbol{\beta}}_\text{Regularization term to reconstruct}$$

Current regularizer form:

$$\begin{aligned}
\mathtt{mean(batch\_dot(x, dot(???, x))) } & \overset{?}{=}
\frac{1}{n} \sum \sum_{\mathrm{batch}} \boldsymbol{\beta} \cdot (\lambda \boldsymbol{S} \cdot \boldsymbol{\beta})
\end{aligned}$$

Making a D and S matrix

```{r gam-Smat}
# make_Dmat <-function(k = 8) {
#   diff(diag(k), differences = 2)
# }
# # S := D^T D
# make_Smat <- function(k = 8) t(make_Dmat(k)) %*% make_Dmat(k)

# Or just take the one from mgcv as I'm presumably supposed to
# Also multiply with lambda for convenience
gam_Smat <- gam_mgcv$sp * gam_mgcv$smooth[[1]]$S[[1]]
```


```{r gam-keras-build-penalized}
smooth_out_reg <- smooth_input %>%
  layer_dense(
    units = 1, activation = "linear", use_bias = FALSE, name = "smooth_effect",
    kernel_regularizer = function(x) {
      k_mean(k_batch_dot(x,
        k_dot(
          tensorflow::tf$constant(gam_Smat, dtype = "float32"),
          x
        ),
      axes = 2
      ))
    }
  )

combined_out_reg <- layer_add(c(linear_out, smooth_out_reg)) %>%
  layer_activation(activation = "sigmoid", name = "output_prob")

# Combine the above to a keras model
gam_nn_reg <- keras_model(
  inputs = c(linear_input, smooth_input),
  outputs = c(combined_out_reg)
)
```

Does it do things?

```{r gam-keras-fit-penalized, cache=TRUE}
gam_nn_reg %>%
  compile(
    loss = "binary_crossentropy", # That's Bernoulli logLik right?
    optimizer = optimizer_rmsprop(),
    metrics = NULL
  )

gam_nn_reg

gam_nn_reg %>%
  fit(
    # First linear input (x2, x3), then smooth input (x1)
    x = list(as.matrix(df[, 3:4]), model.matrix(gam_mgcv)[, 4:12]),
    y = df$y,
    epochs = 200,
    batch_size = 16,
    validation_split = 0.2,
    verbose = TRUE
  )

get_weights(gam_nn_reg)
```

What do the predictions look like?

```{r gam-keras-plots}
gam_nn_reg_preds <- predict(gam_nn_reg, x = list(as.matrix(lpmat[, 2:3]), lpmat[, 4:12]))

test_df %>%
  select(x1) %>%
  mutate(
    unregularized = as.numeric(gam_nn_preds),
    regularized = as.numeric(gam_nn_reg_preds)
  ) %>%
  tidyr::pivot_longer(cols = ends_with("ized")) %>%
  ggplot(aes(x = x1, y = value, color = name)) + 
  scale_color_brewer(palette = "Dark2") +
  geom_path(size = 2) +
  labs(
    title = "Regularized and unregularized smooth effects fit as NN",
    x = "x1", y = "s(x)", color = ""
    ) +
  theme_minimal() +
  theme(legend.position = "top")
```
